Likelihood free inference of the halo model parameters with simulations of galaxy mock catalogs

### 2015-12-18 halo/text/abc.tex
words

\documentclass[12pt, preprint]{aastex}
\usepackage{graphicx}	% For figures
\usepackage{natbib}	% For citep and citep
\usepackage{amsmath}	% for \iint
\usepackage{bbm}
\usepackage[breaklinks]{hyperref}	% for blackboard bold numbers
\usepackage{hyperref}
\hypersetup{colorlinks}
\usepackage{color}
\usepackage{morefloats}
\definecolor{darkred}{rgb}{0.5,0,0}
\definecolor{darkgreen}{rgb}{0,0.5,0}
\definecolor{darkblue}{rgb}{0,0,0.5}
\hypersetup{ colorlinks,
linkcolor=darkblue,
filecolor=darkgreen,
urlcolor=darkred,
citecolor=darkblue }

@@

\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}

@@

\title{Likelihood free inference with measurements of galaxy clustering, group richness, and number denisty }

@@

The Halo Occupation Distribution model (HOD) is a powerful prescription for 
understanding how galaxies populate dark matter halos.
In this work, we use the Approximate Bayesian Computation method, 
ABC, in order to explore the parameters of the halo model with mock
measurements of galaxy number denisty, galaxy clustering, 
and group richness. These observables are forward-modeled in the ABC
by populating halo catalogs in the Bolshoi N-body dark matter simulation 
with the HOD model and making the above measurements on the generated
galaxy catalogs. We use the publicly available software, 
HALOTOOLS to perform the forward-modeling and make the mock measurements.
We find that including group richness in our set of observables leads to stronger
constraints on the HOD parameters that govern the population of satellite galaxies.
Our ABC analysis also successfully constrains HOD parameters for these same
measurements made in SDSS data and our constrained values are consistent with
previous attempts in the literature.

@@

Parameter estimation in cosomlogy is commonly done with Bayesian inference. 
In the Bayesian framework, constraints on cosmological parameters are achieved by 
sampling from the posterior distribution, the probability of the parameters of the 
model given the data. Posteriors are evaluated by multiplying the prior probability 
distribution of the paramters of the model with the likelihood function, the probability 
of the observed data given the parameters of the model.
Therefore, Bayesian inference relies on the evaluation of the likelihood function.

Evaluation of the likelihood requires on making assumption about the functional 
form of the likelihood. It is often assumed that the likelihood is Gaussian distribution 
with a given covariance matrix that needs to be evaluated from simulating a large set of 
mock catalogs, or other techniques such as bootstrap. In order to 
compute the likelihood funciton, one needs to invert a high dimensional covariance 
matrix which makes the likelihood evaluation and therefore sampling from the posterior 
computationally intractable.

@@

\subsubsection{basics}

Talk about generic abc ...

\subsubsection{ABC with adaptive importance sampling}

\section{Data}

@@

>>>>

\documentclass[12pt, preprint]{aastex}
\usepackage{graphicx}	% For figures
\usepackage{natbib}	% For citep and citep
\usepackage{amsmath}	% for \iint
\usepackage{bm}
\usepackage[breaklinks]{hyperref}	% for blackboard bold numbers
\usepackage{hyperref}
\hypersetup{colorlinks}
\usepackage{color}
\usepackage{morefloats}
\definecolor{darkred}{rgb}{0.5,0,0}
\definecolor{darkgreen}{rgb}{0,0.5,0}
\definecolor{darkblue}{rgb}{0,0,0.5}
\hypersetup{ colorlinks,
linkcolor=darkblue,
filecolor=darkgreen,
urlcolor=darkred,
citecolor=darkblue }

@@

\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\la}{\langle}
\newcommand{\ra}{\rangle}
\newcommand{\vep}{\bm{\epsilon}}
\newcommand{\ep}{\epsilon}

@@

\title{Likelihood free inference of the halo model parameters with simulations of galaxy mock catalogs}

@@

The Halo Occupation Distribution (HOD) model is a powerful prescription of how galaxies populate 
dark matter halos that has had much success in reproducing large scale galaxy clustering and understanding
galaxy evolution within the context of the halo model. Constraints on the parameters of the HOD model have 
so far been infered using a likelihood function that assumes, \emph{First} a functional (often times Gaussian) form of the likelihood, 
\emph{Second} analytical model for the galaxy clustering statistics. We use Approximate Bayesian Computation (ABC) 
in order to infer the HOD parameters by assuming a prior over the parameters and a distance function between 
statistical summaries of the data and statistical summaries of the simulations.
Using a mock galaxy catalog created with known HOD parameters, we obtain precise constraints on the HOD parameters that 
reproduces the summary statistics of the mock galaxy catalog accurately. 
Since ABC does not require computing the likelihood, we include observables beyond galaxy clustering such as 
the galaxy number density and the galaxy group richness, for which accurately constructing a likelihood is intractable. 
We demonstrate that adding these measurements to our set of observables leads to stronger constraints on the HOD parameters, 
specifically the ones that govern the satellite galaxy population. Moreover, the constraints we infer that 
reproduce the known HOD parameters illustrate the success of ABC in the context of constraining HOD parameters.

@@

form of the likelihood. It is often assumed that the likelihood is Gaussian distribution 
with a given covariance matrix that needs to be evaluated from simulating a large set of 
mock catalogs, or other techniques such as bootstrap. In order to 
compute the likelihood funciton, one needs to invert a high dimensional covariance 
matrix which makes the likelihood evaluation and therefore sampling from the posterior 
computationally intractable.

@@

ABC is a likelihood free method for inference of a set of parameters that are required for the simulation of a 
physical system, through a forward model, a distance function, and a prior probability distribution over the parameters. 
Let us denote the the data by $\mathcal{D}$. We assume that there exist a forward model $\mathcal{M}$ that 
produces a simulation of the data $\bar{\mathcal{D}}$ for a given set of parameters $\bm{\theta}$
\beq
\bar{\mathcal{D}}_{\theta} = \mathcal{M}(\bm{\theta})
\eeq

Furthermore, ABC requires adopting a distance function between the data $\mathcal{D}$ and the simulation $\bar{\mathcal{D}}_{\theta}$.
The distance function, in general, can be a vector $\bm{d}(\mathcal{D},\bar{\mathcal{D}}_{\theta})$
\begin{eqnarray}
\bm{d}(\mathcal{D},\bar{\mathcal{D}}_{\theta}) &=& [d_{1,\theta}, ... , d_{M,\theta}], \\
d_{i,\theta} &=& d_{i}(\mathcal{S}_{i}(\mathcal{D}), \mathcal{S}_{i}(\bar{\mathcal{D}}_{\theta}))  ~~\mbox{for $i=\{1,...,M\}$},
\end{eqnarray}
where $d_{i}(\dot,\dot)$ is a scalar function that defines a distance 
between $i$-$th$ summary statistics of the data $\mathcal{S}_{i}(\mathcal{D})$ and the same summary statistics of 
the simulation $\mathcal{S}_{i}(\bar{\mathcal{D}}_{\theta})$, and $M$ is the total number of statistical summaries 
used in the analysis.

ABC algorithm draws a large number of samples from the prior distribution over parameter space, 
and rejects every sample for which the distance between the data and the simulation corresponding to that 
sample is not less than a certain threshold $\bm{\epsilon}$. This requirement is satisfied \emph{if and only if}
every component of the distance vector is less than its corresponding threshold. Note that, each summary statistics 
has a separate threshold. This procedure is deone iteratively untill, an investigator is left with a certain number 
of samples that satisfy the distance criteria. The final ensemble approximates the posterior pdf over 
the paramters of the forward model of the data. This sampling method is called \emph{rejected sampling}.

In practice however, rejected sampling is not desirable because $(i)$ it requires fine tuning of the threshold 
$\bm{\epsilon}$ whose elements are small, and $(ii)$ finding a large number of samples that pass the threshold 
criteria is computationally expensive. Alternatively, we draw a large number of proposals from the prior pdf over 
the HOD parameters with the only requirement that they pass the distance criterion. But since we select an 
arbitrarily large initial threhold, all th proposals pass the test. Moreover, we assign simillar weights to all 
the proposals in the initial pool. In the next steps, we draw new proposals from the previos pool through 
weighted sampling, and we perturb the particles by a kernel that is twice the covariance matrix of the proposals 
in the previous pool.

blaalalalal

\subsection{Halo Occupation Distribution}

Halo Occupation distribution is a powerful tool for description of galaxy/halo relation.

\section{Summary statistics}

\section{Forward modeling of the galaxy population with HALOTOOLS}

\section{Mock data}

We use HALOTOOLS to populate halos with galaxies above a certain luminosity threshold $M_r>-20$.

@@

Hello World!
